\section{Introduction}\label{sec:introduction}


Signatures have proven to be fundamental for several applications. The theory is particularly applicable in the algebraic theory of rough paths \cite{geng2021introduction, geng2017reconstruction, fermanian2023new}. This has been shown to be a very interesting way to represent data due to the special properties of time-reversal and the invariance to time reparametrization when representing data via signatures \cite{chevyrev2016primer}. In fact, in a 2013 competition focused on recognizing handwritten Chinese characters, the winning approach represented the characters as arrays based on a "signature" from rough path theory, then classified them using a convolutional neural network \cite{yin2013icdar}. Recently, it has become evident that signatures can be a powerful tool for extracting the geometric shape of any path, even when data is missing.

When the space is discretized and the path is specified by the points it passes through, the resulting representation becomes large and sparse, leading to slow learning. The signature offers a more efficient way to represent a path, helping to mitigate these issues. Importantly, similar smooth curves produce similar signatures. Additionally, the signature is invariant under translations and reparametrizations of the path, making it a robust representation. This, the signature approach represent a non-parametric way of extraction of characteristics features of data.


The goal of this report is to give a short introduction to the importance of signatures and its applications. So following \cite{geng2021introduction} and \cite{chevyrev2016primer}, we divide the report into two parts. In the first part we define
signatures, and then state the fundamental properties that is the invariance of time reparametrization, shuffle product time-reversibility, the Chen's Identity and a short introduction to the log signature. In the second part, we discuss machine
learning application and then the applications signatures to the general rough paths theory.
